---
author: J-O Irisson
output: 
  html_document: 
    css: document.css
---

# Clustering of CC mitigation solutions

*J-G Gattuso, J-O Irisson  
`r Sys.Date()`*


```{r prep, echo=FALSE, message= FALSE}
knitr::opts_chunk$set(
  echo=F,
  comment="",
  cache=T
)
library("tidyverse")
library("stringr")
library("FactoMineR")
library("broom")
library("ggplot2")
library("ggrepel")
library("dendextend")
# devtools::install_github("jiho/chroma")
library("chroma")
```


## Data 

```{r read_data}
## Global effects
t1a <- read_csv(file="table_1a.csv", col_types=cols()) %>% 
        rename(`Alkalinity glob.`=Alkalinity, `Vegetation glob.`=Vegetation)

t1b <- read_csv(file="table_1b.csv", col_types=cols()) %>% 
        rename(`Alkalinity loc.`=Alkalinity, `Vegetation loc.`=Vegetation, `Relocation, Restoration`=Relocation_Restoration)
# score relative and absolute sea level rise together
t1b$Criteria <- str_replace(t1b$Criteria, "Moderate relative sea", "Moderate sea")

t1 <- full_join(t1a, t1b, by="Criteria")

t1$Criteria <- t1$Criteria %>% 
  # abbreviate some words for clarity of plots
  str_replace("until full potential effectiveness", "until full pot. effectiv.") %>% 
  # force order
  factor(levels=unique(.))


## Local effects
t3 <- read_csv(file="table_3.csv", col_types=cols()) 
# make names clearer
t3 <- t3 %>% rename(`Relocation, Restoration`=Relocation_Restoration)
names(t3) <- str_replace(names(t3), "_g", " glob.")
names(t3) <- str_replace(names(t3), "_l", " loc.")
t3$Criteria <- t3$Criteria %>% 
  # make labels clearer
  str_replace("_", " ") %>% 
  str_replace("Warming", "Reduce impact of warming") %>% 
  str_replace("Acidification", "Reduce impact of acidification") %>% 
  str_replace("Sea level rise", "Reduce impact of sea level rise") %>% 
  str_replace("Other drivers", "Reduce impact of other drivers") %>% 
  # force order
  factor(levels=unique(.))

# keep everything
t3a <- t3 %>%
        mutate(Criteria=str_c(Category, Criteria, sep="_")) %>%
        select(-Category)
# summarise per criterion, accross ecosystems
t3c <- t3 %>% select(-Category) %>% group_by(Criteria) %>% summarise_all(mean, na.rm=T)
# summarise per ecosystem, across criteria
t3e <- t3 %>% select(-Criteria) %>% group_by(Category) %>% summarise_all(mean, na.rm=T)


dt <- suppressWarnings(bind_rows(t1, t3a)) %>% gather(key="Solutions", value="Score", -Criteria)
dw <- dt %>% spread(key="Criteria", value="Score") %>% as.data.frame()

dst <- suppressWarnings(bind_rows(t1, t3c)) %>% gather(key="Solutions", value="Score", -Criteria)
# force order of factors
dst$Solutions <- factor(dst$Solutions)                            # alphabetical
dst$Criteria <- factor(dst$Criteria, levels=unique(dst$Criteria)) # current order
dsw <- dst %>% spread(key="Criteria", value="Score") %>% as.data.frame()
```

The data consists of `r ncol(t1)-1` climate-change mitigation solutions

```{r}
levels(dst$Solutions)
```

Each is scored from 1 (sometimes 0) to 5 according to `r nlevels(t1$Criteria)` global criteria

```{r}
levels(t1$Criteria)
```

As well as `r nlevels(t3$Criteria)` local criteria

```{r}
levels(t3$Criteria)
```
evaluated in in `r length(unique(t3$Category))` "ecosystems"

```{r}
sort(unique(t3$Category))
```

For simplicity and/or the requirements of the techniques, the *scores* for these local criteria are often *averaged* across ecosystems.

## Plot of relationships

Fig x: Links between solutions (light gray) and scoring criteria, either global (black) or local (grey). The links are coloured according to the score, from red (bad) to good (green; grey is an undefined score). For each solution, links are ordered by increasing score.

```{r circos, fig.height=9, fig.width=9}
library("circlize")

# prepare data for Chord diagram
d <- dst %>%
  mutate(
    # add a column of ones for the line width
    One=1,
    # replace empty scores
    Score0=replace(Score, is.na(Score), 0),
    # compute colors manually
    Color=ifelse(is.na(Score), "grey", brewer_map(Score0, name="RdYlGn")) # red=0, green=5
  ) %>%
  # sort by Score within each solution
  arrange(Solutions, desc(Score)) %>%
  # reorder columns for chordDiagram
  select(Criteria, Solutions, One, 1:ncol(.)) %>% 
  as.data.frame()

# shorten label
levels(d$Criteria) <- str_wrap(levels(d$Criteria), width=12)
levels(d$Solutions) <- str_wrap(levels(d$Solutions), width=12)

chordDiagram(d,
  # map precomputed color to lines
  col=d$Color,
  order=c(levels(d$Criteria), levels(d$Solutions)),
  # color grid according to type of data: global criterion, local criterion, solution
  grid.col=rep(
    c("grey0", "grey40", "grey80"),
    times=c(nlevels(t1$Criteria), nlevels(t3$Criteria), nlevels(d$Solutions))
  ),
  # remove axis around plot
  annotationTrack=c("name", "grid"),
  # put green lines one top
  link.largest.ontop=TRUE,
  # plot aspect
  h.ratio=0.6, transparency=0.1,
  directional=1, direction.type="diffHeight"
)
```


Fig x: Same legend as above. Boxes should be drawn around solutions and criteria to better define them

```{r alluvial, fig.height=9}
library("alluvial")
# shorten label
levels(d$Criteria) <- str_wrap(levels(d$Criteria), width=16)
levels(d$Solutions) <- str_wrap(levels(d$Solutions), width=16)

alluvial(
  # draw lines, ordered by score in each solution
  d[,2:1], freq=1,
  ordering=list(with(d, order(Solutions, Score)), NULL),
  # map color
  col=d$Color,
  # cleaner aspect
  border=NA, alpha=0.9, xw=0.2, cw=0.12, blocks=F, gap.width=0.15, cex=0.8
)
```


## Clustering

Clustering the full, non-aggregated data set, with hierarchical clustering (Euclidean distance, Ward aggregation method to highlight synthetic groups) suggests three groups:

```{r hclust}
# reformat data for clustering
d <- dw
row.names(d) <- d$Solutions
d <- select(d, -Solutions)
hc <- as.dendrogram(hclust(dist(d), method="ward.D2"))

# cut tree
n <- 3 # based on the genral aspect of the plot

# nice plot
pars <- par(no.readonly=T)
par(mai=c(2,0.5,0.1,0.1))
cols <- hue.colors(n)
hc <- color_labels(hc, k=n, col=cols)
hc <- color_branches(hc, k=n, col=cols)
plot(hc)
par(pars)
```

The same technique for the dataset in which the scores are averaged by ecosystem gives mostly the same grouping

```{r hclust_s}
# reformat data for clustering
d <- dsw
row.names(d) <- d$Solutions
d <- select(d, -Solutions)
hc <- as.dendrogram(hclust(dist(d), method="ward.D2"))

# cut tree
n <- 3 # based on the general aspect of the plot

# nice plot
pars <- par(no.readonly=T)
par(mai=c(2,0.5,0.1,0.1))
cols <- hue.colors(n)
hc <- color_labels(hc, k=n, col=cols)
hc <- color_branches(hc, k=n, col=cols)
plot(hc)
par(pars)
```

The distribution of scores for each criterion in the clusters (computed from the averaged dataset) are

```{r scores, fig.height=10}
# associate groups with the correct solution, preserving the colors compared to the previous plots
groups <- data.frame(
  Solutions=factor(labels(hc)),
  colour=unlist(dendrapply(hc, function(x) {attributes(x)$nodePar$lab.col}))
)
dsg <- left_join(dst, groups, by="Solutions")

# plot
levels(dsg$Criteria) <- str_wrap(levels(dsg$Criteria), 15)
ggplot(dsg) + geom_violin(aes(x=Criteria, y=Score, fill=colour), scale="width", colour=alpha("black", 0.5), na.rm=T) + scale_fill_identity() + coord_flip() + geom_vline(aes(xintercept=8.5))
```

This plot can be read as:

- the green solutions are particularly effective (high scores) against warming, both local and global. They also tend to be more effective against acidification. They are quite ineffective against other drivers and have low scores for global governability, cost effectiveness and co-benefits.
- the red solutions are practical: high readiness, goog governability, few unintended effects, and high co-benefits; however they are not very effective against warming and acidification
- the blue solutions are mostly average everywhere, except they have somewhat good cost effectiveness, low co-benefits, and are not very effective to moderate acidification globally.


## PCA then clustering

A principal component analysis (PCA) can be used to examine the most important correlations and remove noise before the clustering step.

The PCA is performed without scaling the data, assuming that all criteria have equal importance and that those for which the solutions are all scored similarly are not discriminative and should not matter much. For simplicity, we use the scores averaged by ecosystem.

The eigenvalues or the PCA highlights that the first 4 PCs hold significant information

```{r PCA}
# reformat data for PCA
d <- dsw
row.names(d) <- d$Solutions
d <- select(d, -Solutions)

# cleanup labels
names(d) <- str_wrap(names(d), 18)

# perform PCA without scaling (since most variables are in the same range)
pca <- suppressWarnings(PCA(d, graph=F, scale=F))

# inspect eigenvalue
qplot(1:nrow(pca$eig), pca$eig$eigenvalue) + geom_hline(yintercept=mean(pca$eig$eigenvalue), colour="red")
```

Hierarchical clustering (Euclidean distance, Ward aggregation method) is performed on the coordinates of the solutions on those first 4 PCs. As before, 3 groups seem to emerge and are mostly equivalent to those highlighted on the raw data (which is expected of course, since the PCA just summarises the data and removes noise...).

```{r clust_on_PCA}
# perform synthetic clustering
coords <- pca$ind$coord[,1:4]
hc_pca <- as.dendrogram(hclust(dist(coords), method="ward.D2"))

# cut tree
n <- 3 # based on the genral aspect of the plot

# nice plot
pars <- par(no.readonly=T)
par(mai=c(2,0.5,0.1,0.1))
cols <- hue.colors(n)
hc_pca <- color_labels(hc_pca, k=n, col=cols)
hc_pca <- color_branches(hc_pca, k=n, col=cols)
plot(hc_pca)
par(pars)
```

The real interest here is that the clusters can re-projected in the PCA space where the solutions can be individually related to the criteria in which they score well. The two following plots represent:

1. the correlation of criteria based on their scoring, which define the PCA space. The angle between the lines of the criteria reflect their correlation: two lines close to each other means that most solutions have similar scores for these two criteria, two lines at 90ยบ angle means that solutions have scores that are completely uncorrelated for these two criteria.
2. the solutions distributed in the PCA space space according to their scoring patterns and clustered into groups of solutions with similar scoring patterns. If a solution is in the general direction in which a criterion points, it has a high score according to this criterion (and if it is in the opposite direction, it has a low score).

In both cases, the criteria or solutions which are not well represented in this PCA space (i.e. which have quite nondescript scoring patterns) are made semi-transparent.

```{r PCA_and_clust, fig.height=5, fig.width=6}
# plot result of PCA and clustering together

#' Extract data from a PCA object
#'
#' @param x output of FactoMineR::PCA
#' @param data original data; optional
#' @param dimensions PCs to extract
#' @param which objects/rows/individuals or descriptors/columns/variables
augment.PCA <- function(x, data=NULL, dimensions=c(1,2), which=c("ind", "var")) {
  which <- match.arg(which)
  d <- data.frame(
    x[[which]]$coord[,dimensions],
    cos2=rowSums(x[[which]]$cos2[,dimensions]),
    contrib=rowSums(x[[which]]$contrib[,dimensions])
  )
  d$label <- row.names(d)
  if (which == "ind" & !is.null(data)) {
    d <- cbind(d, data)
  }
  return(d)
}

# define groups of solutions from clustering, preserving the colors compared to the dendrogram plot
groups <- data.frame(
  label=labels(hc_pca),
  colour=unlist(dendrapply(hc_pca, function(x) {attributes(x)$nodePar$lab.col}))
)

# extract objects from PCA
pca_i12 <- augment(pca)
# add clusters
pca_i12 <- left_join(pca_i12, groups, by = "label")
# extract variables from PCA
pca_v12 <- augment(pca, which="var")

pvar <- round(pca$eig$`percentage of variance`[1:2], 1)
coord_pca <- list(
  coord_fixed(),
  labs(x=paste0("PC1 (", pvar[1], "%)"), y=paste0("PC2 (", pvar[2], "%)"))
)

pv12 <- ggplot(pca_v12, aes(alpha=cos2)) +
  geom_segment(aes(x=0, y=0, xend=Dim.1, yend=Dim.2), arrow=arrow(angle=15, length=unit(0.02, "npc"))) +
  # geom_text(aes(x=Dim.1, y=Dim.2, label=label), hjust=0.5, vjust=0.5, size=3) +
  geom_text_repel(aes(x=Dim.1, y=Dim.2, label=label), segment.alpha=0.2, min.segment.length=unit(0.02, "npc"), size=3) +
  scale_alpha(limits=c(0,0.9)) +
  scale_x_continuous(breaks=0, expand=c(0.05, 0.2)) +
  scale_y_continuous(breaks=0, expand=c(0.05, 0.2)) +
  coord_pca

pi12 <- ggplot(pca_i12, aes(Dim.1, Dim.2, alpha=cos2, colour=colour)) +
  geom_point(shape=16) + geom_text_repel(aes(label=label), size=3) +
  scale_colour_identity() + scale_alpha(limits=c(0,0.9)) +
  scale_x_continuous(breaks=0) + scale_y_continuous(breaks=0) +
  coord_pca

pv12

pi12
```

These plots are read as follows:

- Solutions effective against warming and acidification are effective at global and local scales (arrows are almost overlapping)
- Solutions with a high degree of readiness are those judged to have to most co-benefits, few unintended impacts and good governability.
- Cost effectiveness, lead time until effectiveness, and effectiveness against sea level rise are all badly projected, meaning that solutions are not very differentiated according to these criteria.
- Mitigation of warming, both global and local, is opposed to governability, mitigating other drivers, co-benefits etc. = most solutions cannot score high on criteria in both of these groups. If a solution is effective for warming, it has poor governability and poor effectiveness on other drivers.
- The solutions effective against acidification have very different levels of readiness, co-benefits, etc. (~90ยบ angle)

and

-  The solutions in the green group are effective against warming and acidification (arrows in the second plot point in their direction in the first plot). Hybrids and Albedo are particularly effective for warming-related stuff. Alkalinity global and Renewables are particularly effective for acidification-related stuff. As already explained, in turn, they have poor scores in Global governability, reducing the impact of other drivers, etc. which point in the opposite direction.
- The solutions in the red group are basically the contrary of those in the green group regarding warming and governability, other drivers, co-benefits, readiness. The solutions Vegetation loc., Pollution, Protection have a high degree of readiness, last long, and have good co-benefits.
- The blue group is not very specific. The only pattern is that Cloud is not effective for mitigating acidification and basically scores bad everywhere.

```{r biplot, fig.height=7, fig.width=7}
# # biplot through vegan
# suppressPackageStartupMessages(library("vegan"))
# dd <- data.frame(lapply(d, function(x) {
#   x[is.na(x)] <- mean(x, na.rm=T)
#   return(x)
# }))
# row.names(dd) <- row.names(d)
# 
# pca_v <- rda(dd, scale=F)
# biplot(pca_v, scaling="sites")
```


To examine the solutions and criteria not well represented in dimension 1 and 2, we can examine dimensions 2 and 3:

```{r PCA_and_clust23, fig.height=5, fig.width=6}
# extract objects from PCA
pca_i23 <- augment(pca, dimensions=c(2,3))
# add clusters
pca_i23 <- left_join(pca_i23, groups, by = "label")
# extract variables from PCA
pca_v23 <- augment(pca, which="var", dimensions=c(2,3))

pvar <- round(pca$eig$`percentage of variance`[2:3], 1)
coord_pca <- list(
  coord_fixed(),
  labs(x=paste0("PC2 (", pvar[1], "%)"), y=paste0("PC3 (", pvar[2], "%)"))
)

ggplot(pca_v23, aes(alpha=cos2)) +
  geom_segment(aes(x=0, y=0, xend=Dim.2, yend=Dim.3), arrow=arrow(angle=15, length=unit(0.02, "npc"))) +
  # geom_text(aes(x=Dim.1, y=Dim.2, label=label), hjust=0.5, vjust=0.5, size=3) +
  geom_text_repel(aes(x=Dim.2, y=Dim.3, label=label), segment.alpha=0.2, min.segment.length=unit(0.02, "npc"), size=3) +
  scale_alpha(limits=c(0,0.9)) +
  scale_x_continuous(breaks=0, expand=c(0.05, 0.2)) +
  scale_y_continuous(breaks=0, expand=c(0.05, 0.2)) +
  coord_pca

ggplot(pca_i23, aes(Dim.2, Dim.3, alpha=cos2, colour=colour)) +
  geom_point() + geom_text_repel(aes(label=label), size=3) +
  scale_colour_identity() + scale_alpha(limits=c(0,0.9)) +
  scale_x_continuous(breaks=0) + scale_y_continuous(breaks=0) +
  coord_pca
```

This does not add much information (as expected since PCs 1 and 2 already capture 62% of the variance). We could still remark that:

- Alkalinity loc is somewhat effective against acidification and not very effective against sea level rise (which seems like it can be expected even by my non-specialist-self ;-) )
- Evolution and Productivity are confirmed to have mostly bad scores in everything.

Hope that helps!
